---
title: "Ukraine NDVI Analysis"
author: "Michael Wellington, Roger Lawes and Petra Kuhnert"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ukraine NDVI Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(stringr)
library(tidyverse)
library(gratia)
library(mgcv)
library(rgdal)
library(sf)
library(mgcViz)
library(scales)
library(gridExtra)
library(ggsci)
library(itsadug)
library(maps)
library(gstat)


library(UkraineCrops)
data(NDVI_mod_df)
data(Ukr_bnds)
```

## Overview
The following script accompanies the paper by Wellington et al. (2022) that explores crop growth, grain exports and fire patterns in Ukraine.  This first vignette focuses on trends in crop growth using NDVI.


## Trend Analysis

# Reading in relevant datasets

We load in NDVI data clipped to Ukraine and load in the relevant shape file. These datasets are also available in the `/data` and `/inst/extdata` directories and can be read in using the `readRDS` and `st_read`  commands respectively.

```{r echo = TRUE, eval = FALSE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

library(UkraineCrops)

data(NDVI_mod_df)
data(Ukr_bnds)

```

# Creating a pre and post-invasion term

The `NDVI_mod_df` includes a war term for pre and post 2022 calendar years but we also need to create a pre and post-invasion term reflected by the date of the invasion (24th of February 2022).

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Allocate factor for invasion
NDVI_mod_df$inv <- as.factor(if_else(NDVI_mod_df$date < '2022-02-24', "0", "1"))
```


# Fitting the Generalised Additive Model (GAM)

We fit a Generalised Additive Model (GAM) (Wood 2006) assuming a Gaussian distribution for the response variable, NDVI and include smooth terms for year, season, space and relevant interactions, which consist of season within year, and space and time. 

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Setting up the GAM structure with key smoothers for trend terms
all_f <- NDVI ~  
  # smooth term for year
  s(year, bs="cr", k=11) + 
  # cyclic term for season
  s(month, bs="cc", k=12) +  
  # smooth term for space
  s(x,y, bs='gp', k=50)  +            
  # seasonal within year
  ti(month, year, bs = c("cc", "cr"), k = c(12,11)) +  
  # space x time
  ti(x, y, year, d = c(2, 1), bs = c("gp", "cr"), 
     k = c(50,11), m=list(2, NA))  
```

We then add in the factor `inv` which allows us to investigate a pre and post invasion effect.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Setting up the GAM structure with an added term reflecting
# the start of the invasion
all_f_inv <- NDVI ~  
  # factor for pre/ post-invasion
  inv + 
  # cyclic term for season
  s(month, bs="cc", k=12, by=inv) +  
  # smooth term for space
  s(x,y, bs='gp', k=50, by=inv) +
  # space x time
  ti(x, y, month, d = c(2, 1), bs = c("gp", "cr"), 
     k = c(50,12), m=list(2, NA), by=inv)  
```

We also explore a GAM structure that fits a term reflecting the pre (2012-21) and post war (2022) periods.  This term which we refer to as `war` represents a binary variable where 1 indicates year 2022 when the war began.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}
all_f_war <- NDVI ~  
  # factor for pre/ post-2022
  war + 
  # cyclic term for season
  s(month, bs="cc", k=12, by=war) +  
  # smooth term for space
  s(x,y, bs='gp', k=50, by=war) +
  # space x time
  ti(x, y, month, d = c(2, 1), bs = c("gp", "cr"), 
     k = c(50,12), m=list(2, NA), by=war) 


```

We now fit each model using the `mgcv` package in `R` and store the results. To accommodate temporal dependencies we fit a correlation term in each model that was the result of an iterative process as outlined in Wood et al. (2017).  The final model fits are presented here.  Each model takes a few minutes to run.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Run GAMs for large datasets
all_gam <- bam(all_f, data=NDVI_mod_df, discrete=TRUE, nthreads=8, rho=0.8)
all_gam_war <- bam(all_f_war, data=NDVI_mod_df, discrete=TRUE, nthreads=8, rho=0.6)
all_gam_inv <- bam(all_f_inv, data=NDVI_mod_df, discrete=TRUE, nthreads=8, rho=0.6)

# Output shows effect of the invasion on cropland NDVI
summary(all_gam_inv)
```

# Model Checking

We examine the fit of the model and check for temporal and spatial autocorrelation in the residuals.  These diagnostics are examined below and sufficiently account for correlation in the data with the given model structure.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Temporal autocorrelation plot
resids <- residuals.gam(all_gam_inv)
acf_plot_ar <- acf(resids, type = "correlation")

plot(acf_plot_ar)

# Variogram for spatial autocorrelation
data_pred <- data.frame(resids = resids, long = NDVI_mod_df$x, lat = NDVI_mod_df$y)
coordinates(data_pred) <- ~long+lat
# Select a sample for compute efficiency  
var_plot <- variogram (resids ~ 1, data = data_pred[sample(1:nrow(data_pred), 10000),])  
plot(var_plot)
```


# Investigating Key Trends

We obtain NDVI predictions for the model and produce maps comparing the baseline period 2012-2021 (top row) with 2022 (bottom row) as per Figure 2 in Wellington et al. (2022). (Note this takes a few minutes to run)


```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}

# Get predictions for plotting
all_gam_war_pred <- predict(all_gam_war, se=TRUE)

all_gam_war_pred_df <- data.frame(pred = all_gam_war_pred$fit,
                                  pred_se = all_gam_war_pred$se.fit,
                                  x = NDVI_mod_df$x,
                                  y= NDVI_mod_df$y,
                                  year = NDVI_mod_df$year,
                                  month = NDVI_mod_df$month,
                                  war=NDVI_mod_df$war)


# Filter to current time period to allow comparison of baseline to 2022
all_gam_war_pred_space_df <- all_gam_war_pred_df %>% filter(month < 7)

# Space plot by month
space_month_year_plot <- ggplot(data=all_gam_war_pred_space_df) + geom_tile(aes(x=x, y=y, fill=pred)) + 
  geom_contour(aes(x=x, y=y, z=pred, col="red"), )  + theme_bw() + facet_grid(war~month) +
  scale_fill_viridis_c(name='NDVI') + coord_equal() +
  labs(x='Longitude', y='Latitude') + guides(alpha="none", colour="none") + geom_sf(data=Ukr_bnds, alpha=0) +
  theme(legend.key.size = unit(1, 'cm'), legend.position = "bottom")

space_month_year_plot



```

We plot the partial effect of year on cropland NDVI in Ukrain with a shaded ribbon denoting plus/minus the standard error from the estimate.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Plot partial effect of year
all_gam_viz <- getViz(all_gam)

all_gam_year_plot_dat <- plot(sm(all_gam_viz, 1))$data$fit

Ukr_year_plot <- ggplot(data = all_gam_year_plot_dat) + 
  geom_line(aes(x = x, y = y)) +
  geom_ribbon(aes(x = x, ymin = y-se, ymax = y+se), alpha = 0.2) + 
  theme_bw() +
  scale_x_continuous(breaks = seq(2012, 2022,1)) +
  labs(y = "NDVI", x = "Year")

Ukr_year_plot


```

# References

Wellington, M., Lawes, R., Kuhnert, P. (2022) Rapid monitoring of crop growth, grain exports, and fire patterns in Ukraine, Nature Food, Under Review.

Wood, S. N., Li, Z., Shaddick, G. & Augustin, N. H. (2017) Generalized Additive Models for Gigadata: Modeling the U.K. Black 166 Smoke Network Daily Data. J. Am. Stat. Assoc. 112, 1199–1210, DOI: 10.1080/01621459.2016.1195744.

Wood, S. N. (2006) Generalized additive models: an introduction with R, Chapman and Hall/CRC.


