---
title: "Ukraine Fire Analysis"
author: "Michael Wellington, Roger Lawes and Petra Kuhnert"
date: "4 August 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ukraine Fire Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(ggplot2)




library(UkraineCrops)
data(fire_mod_df)

```

## Overview
The following script accompanies the paper by Wellington et al. (2022) that explores the fire incidence data at a 25-km spatial resolutions for each day from the 1st of January 2012 through to the 20th of July 2022 in Ukraine.  

We first load the library and relevant fire data.
```{r echo = TRUE, eval = FALSE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

library(UkraineCrops)

data(fire_mod_df)


```

We `tidy` the data to construct the percent fire pixels from total pixels in preparation for plotting.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}

# Note in that dims are 37 x 82, store this so we can calc. % fire pixels from total pixels
pixels <- 37*82

# % fire analysis 
month_pc_ts <- fire_mod_df %>% 
  
  mutate(fire_inc=if_else(fire=="NaN", 0,1)) %>% 
  group_by(year, war,month) %>% 
  summarise(fire_pc_mean = mean(sum(fire_inc)/(pixels*length(unique(date)))))

month_pc_ts <- month_pc_ts %>% group_by(war, month) %>% 
  summarise(fire_pc = mean(fire_pc_mean),
            fire_pc_sd = sd(fire_pc_mean, na.rm = TRUE))

# Cumulative fire area plot
cum_plot_dat <-  fire_mod_df %>% mutate(fire_inc=if_else(fire=="NaN", 0,1)) %>% 
  group_by(year, war,month, day) %>% 
  summarise(fire_pc_mean = mean(sum(fire_inc)/(pixels*length(unique(date))))) %>%
  group_by(war, month) %>% 
  summarise(fire_pc = mean(fire_pc_mean),
            fire_pc_sd = sd(fire_pc_mean, na.rm = TRUE)) %>% 
  mutate(fire_pc_cumsum = cumsum(fire_pc))

```

We produce a plot of the cumulative cropland fire area for the pre invasion (2012-21) and post invasion (2022) periods.  These were calculated as a percentage of pixels where fire was present.  Shaded ribbons in this plot denoste 95% confidence intervals on the smooth of cumulative fire area for both periods investigated.

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

# Figure 3 in manuscript - cumulative fire area
cumsum_fire_plot <- ggplot(data=cum_plot_dat) + geom_smooth(aes(x=month, y=fire_pc_cumsum, col=war, fill=war)) +
  scale_color_npg(name="") + scale_fill_npg(name="") +
  theme_bw() +  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks=seq(1,12,1))+ labs(x="Month", y="Cumulative cropland fire area (%)") 

cumsum_fire_plot
```


``




# References

Wellington, M., Lawes, R., Kuhnert, P. (2022) Rapid monitoring of crop growth, grain exports, and fire patterns in Ukraine, Nature Food, Under Review.



