---
title: "Ukraine Ship Analysis"
author: "Michael Wellington, Roger Lawes and Petra Kuhnert"
date: "25 January 2023"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ukraine Ship Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(sf)
library(scales)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggthemes)
library(ggspatial)
library(ggsci)
library(raster)
library(tidyverse)
library(ggnewscale)
library(gganimate)
library(lubridate)
library(cowplot)


library(UkraineCrops)
data(cargo_year_df)
data(Port_activity_ggdf)

```

## Overview
The following script accompanies the paper by Wellington et al. (2022) that explores the shipping data and in particular, cargo ship routes between 2021 and 2022 for ports around the Ukraine.  

We first load the library and raster data object. This example demonstrates how to generate the map and plots. The raster data has been pre-processed as described in the manuscript. The original data can be downloaded from [EMODNET](https://www.emodnet-humanactivities.eu/view-data.php).
```{r echo = TRUE, eval = FALSE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

library(UkraineCrops)
data(cargo_year_df)
data(Port_activity_ggdf)
```

# Data Curation
We curate a dataset and map constrained to the the major ports in and out of Ukraine.


```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}
sf::sf_use_s2(FALSE)

world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)



world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

world_area <- world %>%  dplyr::filter(name %in% c('Ukraine', 'Russia',
                                                   'Romania', 'Moldova',
                                                   'Bulgaria', 'Turkey', 'Georgia',
                                                   'Greece'))
```

# Map Generation

We now generate the plot the shipping routes through the Black Sea between 2021 and 2022 for January through to August.
```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}
facet_map <- ggplot() + 
  geom_sf(data = world_area, fill="grey", show.legend=FALSE) + 
  coord_sf(xlim = c(25.5, 43.5), ylim = c(40.1, 48.1), expand = FALSE) + 
  xlab('Longitude') + ylab('Latitude') + 
  theme(panel.grid.major = 
          element_line(color = gray(.5), linetype = 'dashed', size = 0.5),
        panel.background = element_rect(fill = 'aliceblue')) +
  new_scale_fill() +
  geom_raster(data = cargo_year_df, aes(x=x, y=y, fill=value)) +
  scale_fill_viridis_c(limits = c(0, 100), oob = scales::squish, 
                       name=expression(paste("Route Density/", km^2)),
                       alpha=0.5) +
  annotate(geom = "text", x = 40.3, y = 45.6, label = "RUSSIA", 
           color = "grey28", size = 4) +
  annotate(geom = "text", x = 33, y = 47.5, label = "UKRAINE", 
           color = "grey28", size = 4) +
  annotate(geom = "text", x = 34, y = 40.7, label = "TURKEY", 
           color = "grey28", size = 3) +
  annotate(geom = "text", x = 28.4, y = 47.4, label = "MOLDOVA", 
           color = "grey28", size = 1.4, angle=-50) +
  annotate(geom = "text", x = 42.4, y = 42.6, label = "GEORGIA", 
           color = "grey28", size = 1.8, angle=-40) +
  annotate(geom = "text", x = 26.9, y = 45.5, label = "ROMANIA", 
           color = "grey28", size = 2.0, angle=-40) +
  annotate(geom = "text", x = 26.7, y = 43.0, label = "BULGARIA", 
           color = "grey28", size = 1.2, angle=-40) +
  annotate(geom = "text", x = 36.5, y = 46.2, label = "Sea of Azov", 
           fontface = "italic", color = "grey22", size = 2.1) +
  annotate(geom = "text", x = 34.3, y = 43.4, label = "Black Sea", 
           fontface = "italic", color = "grey22", size = 4) +
  annotate(geom = "text", x = 37.6, y = 47.1, label = "Mariupol", 
           fontface = "bold", color = "black", size = 3) +
  annotate(geom = "text", x = 37.8, y = 44.7, label = "Novorossiysk", 
           fontface = "bold", color = "black", size = 3) +
  annotate(geom = "text", x = 30.8, y = 46.8, label = "Odesa", 
           fontface = "bold", color = "black", size = 3) +
  annotate(geom = "text", x = 29.5, y = 45.4, label = "Danube Delta", 
           fontface = "bold", color = "black", size = 3) +
  annotate(geom = "text", x = 28.7, y = 44.1, label = "Constanta", 
           fontface = "bold", color = "black", size = 3) +
  #annotation_scale(location = 'bl', width_hint = 0.5, height = unit(0.1, "cm")) + 
  annotation_north_arrow(location = 'tr', which_north = 'true',
                         pad_x = unit(0.05, 'in'), pad_y = unit(0.05, 'in'), 
                         style = north_arrow_fancy_orienteering,
                         height= unit(0.5, "cm"),
                         width= unit(0.5, "cm")) + 
  facet_grid(~year) +
  theme(legend.position="bottom")

```

# Plot Generation

We need to generate the second plot, which shows the cumulative shipping route density at different port centres. Again, we are using pre-processed data in this vignette. Details of pre-processing can be found in Wellington et al. (reference below).

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}
Activity_month_plot <- ggplot(data=Port_activity_ggdf) + geom_line(aes(x=month, y=activity_cumsum, col=year_fac), size=2) +
  facet_wrap(~port, scale="free_y", nrow=3) + theme_bw() + scale_x_continuous(breaks = 1:12, 
        labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')) +
  labs(x="Month", y=expression(paste("Cumulative Route Density/", km^2))) + theme(legend.title=element_blank()) +
  scale_color_manual(values = c("purple", "orange"))

```

# Formatting Final Plot

We can now assemble the plot combining a mapped comparison and cumulative route density.

```{r echo = TRUE, fig.align='center', fig.height=10, fig.width=8, warning=FALSE, messages=FALSE}
Ship_plot <- ggdraw() +
  draw_plot(Activity_month_plot, x=0,y=0, width=1, height=0.5) +
  draw_plot(facet_map, x=0, y=0.5, width=1, height=0.5) +
  theme(plot.background = element_rect(fill="white", color = NA))
Ship_plot

```

# Calculate % change

Finally, we'll calculate the percentage change in shipping activity, year on year to August.

```{r echo = TRUE, fig.align='center', fig.height=10, fig.width=8, warning=FALSE, messages=FALSE}

Port_activity_ggdf %>% filter(month==9, port=='Odesa')
(167.37-353.18)/353.18*100

Port_activity_ggdf %>% filter(month==9, port=='Mariupol')
(134.67-332.55)/332.55*100
```
Interesting that activity at Odesa and Mariupol decreased by 53% and 60%, respectively. This roughly equates to them being closed in 5 of 9 months (5/9*100 = 55.6%). Final calculation for the increase in activity at the Danube Delta below.

```{r echo = TRUE, fig.align='center', fig.height=10, fig.width=8, warning=FALSE, messages=FALSE}
Port_activity_ggdf %>% filter(month==9, port=='Danube Delta')
(94.68-55.37)/94.68*100

```

# References

Wellington, M., Lawes, R., Kuhnert, P. (2022) Rapid monitoring of crop growth, grain exports, and fire patterns in Ukraine.



