---
title: "Ukraine Ship Analysis"
author: "Michael Wellington, Roger Lawes and Petra Kuhnert"
date: "4 August 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ukraine NDVI Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(sf)
library(scales)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggthemes)
library(ggspatial)
library(ggforce)
library(raster)
library(ggnewscale)
library(gganimate)
library(gifski)


library(UkraineCrops)
data(fire_mod_df)

```

## Overview
The following script accompanies the paper by Wellington et al. (2022) that explores the shipping data and in particular, cargo ship routes between 2021 and 2022 for ports around the Ukraine.  

We first load the library and relevant shipping data.
```{r echo = TRUE, eval = FALSE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE}

library(UkraineCrops)

data(cargo_month_df)


```
We curate a dataset and map constrained to the the major ports in and out of Ukraine.


```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}
sf::sf_use_s2(FALSE)

world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)



world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

world_area <- world %>%  dplyr::filter(name %in% c('Ukraine', 'Russia',
                                                   'Romania', 'Moldova',
                                                   'Bulgaria', 'Turkey', 'Georgia',
                                                   'Greece'))

cargo_month_df$date <- as.Date(paste("01", cargo_month_df$name), "%d %b.%Y")

ships_war_df <- cargo_month_df %>% 
  mutate_at(vars(date), dplyr::funs(year = lubridate::year, month = lubridate::month,
                                    day =lubridate::day)) %>% 
  filter(year > 2020, month < 7)

ships_war_df$month_b <- format(ships_war_df$date, '%b')
ships_war_df$month_b <- factor(ships_war_df$month_b, 
                                            levels = c("Jan", "Feb", "Mar",
                                                       "Apr", "May", "Jun"))

```

We now plot the shipping routes through the Black Sea between 2021 and 2022 for January through to June. This represents Figure 4 of Wellington et al. (2022)

```{r echo = TRUE, fig.align='center', fig.height=5, fig.width=8, fig.asp=.6, warning=FALSE, messages=FALSE}

facet_map_vert <- ggplot() + 
  geom_sf(data = world_area, fill="grey", show.legend=FALSE) + 
  coord_sf(xlim = c(25.5, 43.5), ylim = c(40.1, 48.1), expand = FALSE) + 
  xlab('Longitude') + ylab('Latitude') + 
  theme(panel.grid.major = 
          element_line(color = gray(.5), linetype = 'dashed', size = 0.5),
        panel.background = element_rect(fill = 'aliceblue')) +
  new_scale_fill() +
  geom_raster(data = ships_war_df, aes(x=x, y=y, fill=value)) +
  scale_fill_viridis_c(limits = c(0, 100), oob = scales::squish, 
                       name=expression(paste("Route Density/", km^2, "/month")),
                       alpha=0.5) +
  annotate(geom = "text", x = 40.3, y = 45.6, label = "RUSSIA", 
           color = "grey28", size = 3) +
  annotate(geom = "text", x = 33, y = 47.5, label = "UKRAINE", 
           color = "grey28", size = 3) +
  annotate(geom = "text", x = 34, y = 40.7, label = "TURKEY", 
           color = "grey28", size = 3) +
  annotate(geom = "text", x = 28.4, y = 47.4, label = "MOLDOVA", 
           color = "grey28", size = 1.2, angle=-50) +
  annotate(geom = "text", x = 42.4, y = 42.6, label = "GEORGIA", 
           color = "grey28", size = 1.4, angle=-40) +
  annotate(geom = "text", x = 26.9, y = 45.5, label = "ROMANIA", 
           color = "grey28", size = 1.7, angle=-40) +
  annotate(geom = "text", x = 26.7, y = 43.0, label = "BULGARIA", 
           color = "grey28", size = 1.1, angle=-40) +
  annotate(geom = "text", x = 36.5, y = 46.2, label = "Sea of Azov", 
           fontface = "italic", color = "grey22", size = 1.3) +
  annotate(geom = "text", x = 34.3, y = 43.4, label = "Black Sea", 
           fontface = "italic", color = "grey22", size = 3) +
  annotate(geom = "text", x = 37.6, y = 47.1, label = "Mariupol", 
           fontface = "bold", color = "black", size = 2) +
  annotate(geom = "text", x = 37.8, y = 44.7, label = "Novorossiysk", 
           fontface = "bold", color = "black", size = 2) +
  annotate(geom = "text", x = 30.8, y = 46.5, label = "Odesa", 
           fontface = "bold", color = "black", size = 2) +
  annotate(geom = "text", x = 28.7, y = 44.1, label = "Constanta", 
           fontface = "bold", color = "black", size = 2) +
  annotation_north_arrow(location = 'tr', which_north = 'true',
                         pad_x = unit(0.05, 'in'), pad_y = unit(0.05, 'in'), 
                         style = north_arrow_fancy_orienteering,
                         height= unit(0.5, "cm"),
                         width= unit(0.5, "cm")) + 
  facet_grid(month_b~year) +
  theme(legend.position="bottom")

facet_map_vert


```


# References

Wellington, M., Lawes, R., Kuhnert, P. (2022) Rapid monitoring of crop growth, grain exports, and fire patterns in Ukraine, Nature Food, Under Review.



